---
title: "eda"
author: "Anni and Daniel"
format: html
editor: visual
execute:
  warning: false
  message: false
  echo: true
---

```{r}
library(tidyverse)   
library(skimr)       
library(knitr)      
library(kableExtra)  
library(scales)      
```

```{r}
setwd("/Users/annieyu/math244_project")
wage_data <- read_csv("data/wage_data_raw.csv")
```

### Exploring the Data

dollar difference between male and female median earnings. This variable captures the raw income disparity and allows us to examine the magnitude of the earnings gap. We expect this value to be generally positive, indicating men have a higher median earnings across most occupations.

A second outcome variable we are using, wage_percent_of_male, is included in our dataset. It measures female median earnings as a percentage of male median earnings. A value of 100 would indicate perfect parity; values below 100 indicate women earn less than men. This variable allows us to analyze relative disparity. We expect this variable to generally fall below 100 for most occupations.

Our key explanatory variables are:

-percent_female: The percent of females for specific occupation

-total_workers: Total estimated full-time workers \> 16 years old

-major_category: Broad category of occupation

-minor_category: Fine category of occupation

-occupation: Specific job/career

-year: years from 2013-2016

```{r}
#Select the variables we plan to use
wages_selected <- wage_data %>%
  select(wage_percent_of_male, percent_female, 
        year, major_category, minor_category,
        percent_female, total_workers, 
        major_category, minor_category, 
        occupation, year)
```

```{r}
# Summary table of numeric variables (except year)
wages_selected %>%
  select(where(is.numeric), -year) %>%
  summarise(across(everything(),
    list(
      N = ~sum(!is.na(.)),
      Mean = ~round(mean(., na.rm = TRUE), 1),
      SD = ~round(sd(., na.rm = TRUE), 1),
      Min = ~round(min(., na.rm = TRUE), 1),
      Median = ~round(median(., na.rm = TRUE), 1),
      Max = ~round(max(., na.rm = TRUE), 1)
    )
  )) %>%
  pivot_longer(
    everything(),
    names_to = c("Variable", "Statistic"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  pivot_wider(names_from = Statistic, values_from = value) %>%
  kable(caption = "Table 1: Summary Statistics for Key Variables") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

### Data Wrangling and Transformation

```{r}
#Create a new variable: wage_gap
wage_data <- wage_data %>%
  mutate(wage_gap = total_earnings_male - total_earnings_female)
    
```

```{r}
# Cleaned dataset version 1: Dropped occupations with any missing wage_percent_of_male values

wage_data_complete_cases <- wage_data %>%
  group_by(occupation) %>%
  filter(!any(is.na(wage_percent_of_male))) %>%
  ungroup()

setwd("/Users/annieyu/math244_project")
write_csv(wage_data_complete_cases, "data/wage_data_complete_cases.csv")
```

```{r}
# Cleaned dataset version 2: Retained all observations, computed missing wage_percent_of_male vlaues

wage_data_imputed <- wage_data %>%
  mutate(
    wage_percent_of_male = ifelse(
      is.na(wage_percent_of_male),
      (total_earnings_female / total_earnings_male) * 100,
      wage_percent_of_male
    )
  )

setwd("/Users/annieyu/math244_project")
# Save dataset
write_csv(wage_data_imputed, "data/wage_data_imputed.csv")
```

We saw a considerable amount of missing values in the wage_percent_of_male column of our dataset. The author of TidyTuesday noted this is because those occupations have small sample sizes in total_workers, workers_male, and workers_female. The wage_percent_of_male metric calculated from those values may be unrepresentative and was thus reported as NA.

To address this, we conducted two separate analyses. In one version, we drop all observation with NA wage_percent_of_male values, and also filter out all observation of that same occupation across other years for better reliability and consistency for our time-series analysis.

In the other, we retained these observations, as occupations with small workforce sizes or extreme gender composition may actually capture insightful patterns. For this set of analysis, we recomputed the NA wage_percent_of_male values using the total_earnings_male and total_earnings_female columns.

In terms of data transformation, we created a new variable: wage_gap. It measures the dollar difference between the male and female median earnings (male minus female). This variable captures the magnitude of the wage disparity in absolute terms and complements the percentage-based measure.

### Data Visualization

```{r}
# Plot 1: Distribution of wage_percent_of_male (Complete-Cases)

ggplot(wage_data_complete_cases, aes(x = wage_percent_of_male)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 40,
                 fill = "#4472C4",
                 color = "white",
                 alpha = 0.85) +
  geom_density(color = "#C00000", linewidth = 1.2) +
  geom_vline(xintercept = 100, linetype = "dashed",
             color = "black", linewidth = 0.8) +
  geom_vline(xintercept = median(wage_data_complete_cases$wage_percent_of_male, na.rm = TRUE),
             color = "#C00000", linewidth = 0.8) +
  scale_x_continuous(limits = c(0, 150)) +
  scale_y_continuous(limits = c(0, 0.06)) +
  labs(
    title = "Distribution of Wage Ratio Across Occupations (Complete-Cases)",
    x = "Female Earnings as % of Male Earnings",
    y = "Density",
    caption = "Source: BLS via TidyTuesday | 2013–2016"
  ) +
  theme_minimal(base_size = 13)
  
  
setwd("/Users/annieyu/math244_project")

# Save the figure
ggsave("figures/plot1_wage_ratio_distribution_complete_cases.png", width = 8, height = 5, dpi = 300)
```

```{r}
# Plot 2: Distribution of wage_percent_of_male (All Occupations)

ggplot(wage_data_imputed, aes(x = wage_percent_of_male)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 40,
                 fill = "#4472C4",
                 color = "white",
                 alpha = 0.85) +
  geom_density(color = "#C00000", linewidth = 1.2) +
  geom_vline(xintercept = 100, linetype = "dashed",
             color = "black", linewidth = 0.8) +
  geom_vline(xintercept = median(wage_data_imputed$wage_percent_of_male, na.rm = TRUE),
             color = "#C00000", linewidth = 0.8) +
  scale_x_continuous(limits = c(0, 150)) +
  scale_y_continuous(limits = c(0, 0.06)) +
  labs(
    title = "Distribution of Wage Ratio Across Occupations (All Occupations)",
    x = "Female Earnings as % of Male Earnings",
    y = "Density",
    caption = "Source: BLS via TidyTuesday | 2013–2016 "
  ) +
  theme_minimal(base_size = 13)

setwd("/Users/annieyu/math244_project")

# Save the figure
ggsave("figures/plot2_wage_ratio_distribution_imputed.png", width = 8, height = 5, dpi = 300)
```

```{r}
# Plot 3: Trend of median wage parity over time, by major_category (Complete-Cases)
# Outcome variable (wage_percent_of_male) x time x category

wage_data_complete_cases %>%
  group_by(year, major_category) %>%
  summarise(median_wage_pct = median(wage_percent_of_male, na.rm = TRUE),
            .groups = "drop") %>%
  ggplot(aes(x = year, y = median_wage_pct, 
             color = major_category, group = major_category)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 100, linetype = "dashed", 
             color = "black", alpha = 0.5) +
  scale_x_continuous(breaks = 2013:2016) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     limits = c(70, 110)) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title    = "Wage Ratio by Industry Sector (Complete-Cases)",
    subtitle = "Dashed line = wage parity (100%)",
    x        = "Year",
    y        = "Median Female Earnings as % of Male",
    color    = "Industry Sector",
    caption  = "Source: BLS via TidyTuesday | 2013–2016"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40"),
    legend.position = "right",
    legend.text   = element_text(size = 9)
  )

setwd("/Users/annieyu/math244_project")

# Save the figure
ggsave("figures/plot3_wage_ratio_trend_by_sector_complete_cases.png", width = 10, height = 6, dpi = 300)
```

```{r}
# Plot 4: Trend of median wage parity over time, by major_category (All Occupations)
# Outcome variable (wage_percent_of_male) x time x category

wage_data_imputed %>%
  group_by(year, major_category) %>%
  summarise(median_wage_pct = median(wage_percent_of_male, na.rm = TRUE),
            .groups = "drop") %>%
  ggplot(aes(x = year, y = median_wage_pct, 
             color = major_category, group = major_category)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 100, linetype = "dashed", 
             color = "black", alpha = 0.5) +
  scale_x_continuous(breaks = 2013:2016) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     limits = c(70, 110)) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title    = "Wage Ratio by Industry Sector (All Occupations)",
    subtitle = "Dashed line = wage parity (100%)",
    x        = "Year",
    y        = "Median Female Earnings as % of Male",
    color    = "Industry Sector",
    caption  = "Source: BLS via TidyTuesday | 2013–2016"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40"),
    legend.position = "right",
    legend.text   = element_text(size = 9)
  )

setwd("/Users/annieyu/math244_project")

# Save the figure
ggsave("figures/plot4_wage_ratio_trend_by_sector_imputed.png", width = 10, height = 6, dpi = 300)
```
