---
title: "eda"
author: "Anni and Daniel"
format: html
editor: visual
execute:
  warning: false
  message: false
  echo: true
---

```{r}
library(tidyverse)   
library(skimr)       
library(knitr)      
library(kableExtra)  
library(scales)      
```

```{r}
setwd("/Users/annieyu/math244_project")
wage_data <- read_csv("data/jobs_gender_raw.csv")
```

### Exploring the Data

dollar difference between male and female median earnings. This variable captures the raw income disparity and allows us to examine the magnitude of the earnings gap. We expect this value to be generally positive, indicating men have a higher median earnings across most occupations.

A second outcome variable we are using, wage_percent_of_male, is included in our dataset. It measures female median earnings as a percentage of male median earnings. A value of 100 would indicate perfect parity; values below 100 indicate women earn less than men. This variable allows us to analyze relative disparity. We expect this variable to generally fall below 100 for most occupations.

Our key explanatory variables are:

-percent_female: The percent of females for specific occupation

-total_workers: Total estimated full-time workers \> 16 years old

-major_category: Broad category of occupation

-minor_category: Fine category of occupation

-occupation: Specific job/career

-year: years from 2013-2016

```{r}
#Select the variables we plan to use
wages_selected <- wage_data %>%
  select(wage_percent_of_male, percent_female, 
        year, major_category, minor_category,
        percent_female, total_workers, 
        major_category, minor_category, 
        occupation, year)
```

```{r}
# Summary table of numeric variables (except year)
wages_selected %>%
  select(where(is.numeric), -year) %>%
  summarise(across(everything(),
    list(
      N = ~sum(!is.na(.)),
      Mean = ~round(mean(., na.rm = TRUE), 1),
      SD = ~round(sd(., na.rm = TRUE), 1),
      Min = ~round(min(., na.rm = TRUE), 1),
      Median = ~round(median(., na.rm = TRUE), 1),
      Max = ~round(max(., na.rm = TRUE), 1)
    )
  )) %>%
  pivot_longer(
    everything(),
    names_to = c("Variable", "Statistic"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  pivot_wider(names_from = Statistic, values_from = value) %>%
  kable(caption = "Table 1: Summary Statistics for Key Variables") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

### Data Wrangling and Transformation
